When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in the Contemporary Press

### 2020-07-01 code/Submissionv1.Rmd
made version changes

---
title: "When the Pen Becomes a Sword: Retired Military Officers and Opinion-Editorial Usage in the US"
author: "anonymous"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper seeks to explain how retired military officers have attempted to shape US public opinion through the publication of op-ed pieces in major newspapers. I construct and analysze a data set that captures more than 300 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1975, and ultimately find two trends that stand out. The first is that retired military officers are increasingly writing about topics that are outside of traditional security related topics. The second trend is that retired military officers are increasingly criticizing Presidential administrations, especially that of President Donald Trump, for reasons other than disputes regarding international uses of force. Given the polarizing political climate in the United States and that a Presidential election will occur this fall, these two trends are concerning from the perspective of civil-military relations and suggest that the norm of a professional, apolitical military establishment is under significant strain._

@@

On June 1 2020, the President of the United States, Donald Trump, after giving remarks at the Rose Garden of the White House in Washington D.C., walked to St. John's Church after security forces visibly displaced throngs of protestors in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and after several minutes, returned to the White House.

In the aftermath of what the American mainstream media alleged had amounted to nothing more than a publicity stunt, several members of the President's cabinet members and advisors sought to distance themselves from the President's actions, and more broadly, the President's threat of invoking the Insurrection Act of 1807 to use active-duty forces to quell domestic rioting. The Secretary of Defense, Dr. Mark Esper, stated that he "did not support" use of the 1807 Insurrection Act to deploy active duty military to establish order, and General Mark Milley, the Chairman of the Joint Chiefs of Staff, later told a graduating class from the National Defense University that he "should not have been [joining the President on the walk to St. John's Church]," and that his presence "created a perception of the military involved in domestic politics" [@esper_secretary_2020; @milley_official_2020].

Multiple retired American generals also took to the press to denounce the Trump administration's responce to ongoing protests and riots occurring the aftermath of George Floyd's death in Minneapolis. But for those paying close attention to these retired military elites, it is clear that not all who spoke out did so uniformly. Across a spectrum, there were some retired officers who acknowledged that what had occurred was, at least in a general sense, negative in terms of its impact on civil military relations, while there were also several other officers who went farther in their criticism, directly blasting the President or members of his cabinet for actions taken.

This paper examines the way in which a narrow subset of elites - retired military officers - has attempted to influence public opinion in contemporary America through the use of opinion-editorial pieces. It is an interesting question worthy of examination for several reasons and from a variety of perspectives. First, even after leaving active duty service, military officers remain part of an institution - the military - that maintains considerable trust among the American public. A 2018 Pew Research Center poll determined that 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. Therefore, when leaders of the most trusted institution in the country speak, the public seems to trust them to a high degree. Second, it is an interesting question because as an institution, the military prides itself on being apolitical. In an era of increasing political polarization [@layman_party_2006; @hare_polarization_2014], examining the extent to which opinion-editorial usage by retired military officers has changed may reveal one way in which the military institution has altered its behavior in a polarized political environment, and especially, in the months leading up to the Presidential Election in November of this year.

This paper proceeds in three parts. First, I briefly review the basis upon which the norm of apolitical behavior of the military as an institution is based, as well as describe how a polarized environment may impact these norms. Here I review some of the academic literature on civil-military relations and public opinion, as well as what current military doctrine states on the topic. Second, I describe the analytical approach and the data used in this project. Assembling an original data set that includes more than 300 opinion pieces published by retired military officers in major American newspapers since 1975, I conduct a mixed-methods approach which ultimately reveals two important findings: the first is that although the vast majority of opinion pieces published in major newspapers seem to reflect the military as a partisan institution, retired military officers are increasingly writing about topics that are not immediately related to national security, and second, retired military officers are criticizing if not altogether attacking political administrations more often than they did in previous eras. Third, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a highly polarized political climate.

# The Military as an Apolitical Institution

As a norm, the military acting as an objective, apolitical institution is deep and storied in the United States. In a thriving democracy such as the US, an apolitical military can best serve the Republic by defending it regardless of which political party of group holds power. One of the strongest origins of the norm of the military as an apolitical institution stems from the work of the political scientist, Samuel Huntington. Written in 1957 at the height of the Cold War, his book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced [@huntington_soldier_1957]. The second force was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society, and that these forces were at odds with each other. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution ensured that the military remained "professional," and thus, untainted by political interests and concerns. In this way, Huntington posited that the American military would be up to the task of defeating the very real threats it would face while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant today, came from S.E. Finer, whose 1962 book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962, 22]. Finer's main concern was that of military professionals becoming so professional that they develop the idea that they somehow know better than their elected civilian leadership, particularly on thorny and controversial concerns. Indeed, Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. More important than viewing themselves as professionals, Finer argued that what the military needed in order to stay out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the concept are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situation should be rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military have been cultivated. The operational military - those who prepare for fighting and dying on the battlefield - have also been central in upholding the military's non-partisan traditions. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific who is also well known for the controversy surrounding his conduct in the Korean War and his eventual dismissal by President Truman, told the cadets at West Point in May of 1962.

@@

Current military doctrine and practices likewise reinforce the notion of the military remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General (Retired) Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. General Martin Dempsey, in an opinion piece published in the Washington Post on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although the notion of an apolitical military is the theoretical goal of the institution, the practical record remains far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory from time to time. What we have established, though, is clear: a non-partisan military that performs its military mission well has long been the goal of both civil-military relations theorists and what enables the American public to put trust and confidence in the military as an institution.

# The Media, Public Opinion, and Polarization

Scholars have long examined the role of elites in impacting public opinion. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion were the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller believes that throughout modern day America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information. He is rather acknowledging that most citizens care most about the world that is not "beyond their personal experience," that is, the lives which they must interact with each and every day [@zaller_nature_1992]. His point is that elites play a significant role in helping to form and sway public opinion.

Michael Robinson further developed an argument in which he examines how elite cues are received and processed. Robinson argues that partisan groups, as a result of ongoing political polarization, are cognitively biased in distinct ways in how they receive and process information [@robinson_who_2019]. As a result, Robinson found that partisans will express confidence in the military in vastly differing ways even after neutral, non-partisan information about various types of military failure is presented to these groups [@robinson_who_2019]. If Republicans are less likely to express a reduction in confidence in the military as an institution than Democrats are, as Robinson alleges, then even accurate, unbiased information is potentially irrelevant, as these differing partisan groups are pre-programmed to selectively see a particular situation as they wish to see it [@robinson_who_2019]. His main point serves to warn us regarding confidence in the military as an institution: there may be real reasons to lose confidence in the military as a result of a particular incident, but partisan filters will prevent a reduction of confidence from actually being expressed.

Zaller's and Robinson's works are important considerations in synchronizing the relevant literature for this project. Zaller's framework makes it clear that it is a perfectly reasonable assumption that retired military officers who choose to express their opinions through the form of opinion-editorial pieces constitutes a form of elite discourse. Retired military officers have at least twenty years of experience, and in the case of multi-star generals or admirals, usually close to 35 or 40 years of service. Indeed, most retired senior generals and admirals have spent their professional careers solely in the military, and have extensive experience while serving in leadership positions of large organizations and, in many cases, during times of war. Retired military officers are thus elites in the field of national security, and according to Zaller, they rightfully have a potentially significant contribution to make with respect to impacting public opinion in these areas. And yet, we also know that the American public will not necessarily evaluate the opinions of these military experts objectively, but rather through pre-conceived and potentially partisan lenses [@robinson_who_2019]. This consideration will be helpful when we consider our findings and potential implications.

# Hypotheses

To test whether there has been variation in opinion-editorial usage by retired military officers in the United States, I will examine opinion-editorial pieces published in the five leading newspapers of the United States since January 1, 1979: _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. This date range is important, interesting, and rich, as it captures a multitude of changes and debates that have involved the military in this time period, all of which are captured in the dataset constructed for this project.

Consider first that in terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, and a host of smaller but important military efforts in places such as Somalia, Haiti, and Grenada. Each of these conflicts and wars have themselves varied significantly, including in intensity and and their conclusivity.

Second, the military has undergone significant social and institutional changes in this time period. In the late 1970s, the military had ended the draft only somewhat recently (1973). The data includes discussion of institutional culture that rocked the Navy following the Tailhook Scandalin 1991, as well as debate over first instituting and ultimately repealing the "Don't Ask, Don't Tell" Policy. The data also includes discussions of changes occurring in the Obama Administration, such as the opening of military occupational specialties to women, and frustrations expressed by officers at much of President Trump's foreign policy.

Third, the media itself has undergone significant changes since the late 1970s. Since the mid-2000s, social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, but overall circulation is still down today compared to earlier years, even after factoring in expanse of digital options available today. As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million and 54 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million.

Given these significant changes involving the American military over the past forty years, and the substantial changes that have impacted the media and the manner in which Americans consume information, several specific observations are worth noting about opinion-editorial pieces. The first is that opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. They are different from interviews or commentary that may be asked because the author must take time to put forth his or her position in writing. Second, publishing opinion pieces, especially for major newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by figures that include heads of state, members of congress, governmental figures, and leaders of industry and academia.

Attempting to reconcile the facts that the military prides itself on being a non-partisan and apolitical institution, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 40 years, and that the US political environment is increasingly polarized along partisan lines, I adopt the following hypotheses:

_$H_1$: The majority of op-eds published by retired military officers will be apolitical and non-partisan in tone, but there will be an increase in the number of op-eds criticizing civilian administrations in the past several years, as polarization has dramatically worsened._

_$H_2$: Because overall newspaper consumption has dropped in the past 40 years, there will be fewer opinion pieces over time authored by military officers._

# The Data

To test my hyptheses, I first create a data set that includes all identifiable opinion-editorial pieces authored by retired military officers from January 1, 1979 through June 30, 2020. Using the Factiva and Nexus-Uni databases, I filter a search that utilizes appropriate parameters and uses specific keywords in the byline (author) field such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "general," "admiral," and so forth. I also include in my search both print and online offerings to ensure that I am capturing any opinion pieces published digitally.

I further notice that when collecting and reviewing the data, some authors are identified as military officers in some opinion pieces, but not in others. This was especially true regarding Andrew Bacevich, a retired Army Colonel who then became a prominent historian and author.^[
This happened with several other authors, such as Air Force General (Ret) Michael Hayden, who also served as the Director of the National Security Agency. Where I could identify these authors as military officers, I included the article and did a separate search under the author's last name to see if there were any missing articles that were not caught in my search. 
] I decide to conduct an additional search for Andrew Bacevich and include any other opinion pieces that were published in the sources and over the date range I am examining where he was not identified as a retired military officer. My basis for this decision is that there is a reasonable expectation that some readers would know his background as a retired military officer since it had been previously self-divulged in previous opinion publications.

Ultimately, the data set ends up containing 216 opinion-editorial publications, penned by 129 different authors or author teams.^[There are several co-authored pieces.  Where General (Ret.) David Petraeus wrote one publication and another one was written by General (Ret.) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] Additionally, 10 pieces are written unanimously, with nods by the authors within the article to their being retired from the military in some capacity.

# Methodological Approach and First Level-Analysis

To test my hypotheses, I conduct a mixed-methods approach that attempts to look for evident trends and patterns among the data. I plan to conduct two steps of analysis.  The first is to categorize each observation by subject, and the second is to categorize each observation in terms of its deviation from traditional civil-military norms of the military as remaining an apolitical institution.

In the first level of analysis, after capturing key biographical and publication information, such as author name, article title, source, publication date, and branch of service and rank of the author, I code each observation into one of several binary categories that corresponds to the overall subject of the publication. The goal is to create a coding scheme by subject that is sufficiently broad but also specific enough that is helpful in differentiating the overall main point of each publication. The following coding scheme is therefore devised and implemented (all variables are binary, taking on the value of 1 if the publication corresponds to the general definition of the variable, and 0 otherwise):

>>>>

---
title: "When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in the Contemporary Press"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |
  _This paper seeks to explain how retired military officers have attempted to shape US public opinion through the publication of op-ed pieces in major newspapers. I construct and analyze a data set that captures more than 300 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1975, and ultimately find two trends that stand out. The first is that retired military officers are increasingly writing about topics that are outside of traditional security related topics. The second trend is that retired military officers are increasingly criticizing Presidential administrations, especially that of President Donald Trump, for reasons other than disputes regarding international uses of force. Given the polarizing political climate in the United States and that a Presidential election will occur this fall, these two trends are concerning from the perspective of civil-military relations and suggest that the norm of a professional, apolitical military establishment is under significant strain._

@@

On June 1 2020, the President of the United States, Donald Trump, after giving remarks at the Rose Garden of the White House in Washington D.C., walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and after several minutes, returned to the White House.

In the days and weeks that followed, a full blown civil-military relations ensued [@feaver_military_2020]. Several cabinet members and uniformed military leaders sought to distance themselves from the President's threat of invoking the Insurrection Act of 1807 to use active-duty forces to quell domestic riots that had initiated following the death of George Floyd in Minneapolis. The Secretary of Defense, Mark Esper, stated that he "did not support" use of the 1807 Insurrection Act to deploy active duty military to establish order, and General Mark Milley, the Chairman of the Joint Chiefs of Staff, told a graduating class from the National Defense University that he "should not have been [joining the President on the walk to St. John's Church]," and that his presence "created a perception of the military involved in domestic politics" [@esper_secretary_2020; @milley_official_2020]. Multiple retired American generals also spoke out against the Trump administration's response  But for those paying close attention to what these retired military officers said, it is clear that there was some noticeable variation in their remarks. While all of them expressed deep concern over the way in which the President was threatening to use the military [@brooks_dismay_2020; @dempsey_former_2020 ], some attacked the President in much more forceful and personal terms [@goldberg_james_2020; @mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through the use of opinion-editorial pieces. It is an interesting question worthy of examination for at least three reasons. First, even after leaving active duty service, military officers remain part of the military institution - one that maintains considerable trust and credibility among the American public. A 2018 Pew Research Center poll determined that 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. The second reason this question is important is because as an institution, the military prides itself on being nonpartisan and apolitical, an ideal that is undoubtedly under significant stress in an era of increased political polarization [@layman_party_2006; @hare_polarization_2014]. Third, this question is worth exploring because it may reveal the degree to which the first two factors - military credibility and political polarization - have influenced the concept and norm of professionalism among the military as an institution. All of these factors are worthy of exploration, especially ahead of what promises to be a very spirited presidential election this coming fall.

This paper proceeds in three parts. First, I briefly review the basis upon which the norm of apolitical and nonpartisan behavior of the military as an institution is based. Here I briefly establish a basis for political polarization in contemporary America as well as the high degree of credibility the military as an institution presently enjoys. Second, I describe the analytical approach and the data used in this project. Assembling an original data set that includes more than 300 opinion pieces published by retired military officers in major American newspapers since the late 1970s, I conduct a mixed-methods approach which ultimately reveals that although the vast majority of opinion pieces published in major newspapers seem to uphold the ideal of the military as apolitical and non-partisan, this trend has been weakening in recent years. The data shows that retired military officers are increasingly writing about topics that are not immediately related to national security, as well as increasingly criticizing if not altogether attacking political administrations more often than they did in previous eras. Third, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a highly polarized political climate. 
\singlespace
# The Non-Partisan Military Institution, Credibility, and Polarization  
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is deep. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington. In his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, Huntington asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced [@huntington_soldier_1957]. The second force was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society, and that these forces were at odds with each other. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given today's polarizing environment, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer's concern was that of military officers adopting an ethic of professionalism such that they would actually be more likely - not less likely - to become involved in political issues and concerns [@finer_man_1962, 22-23]. Finer was especially concerned that the adoption of professionalism alone would lead to military professionals believing that they knew what was best for their country, over and above that of the specific, elected civilian leaders in office. Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. Finer's suggestion was that because of the perils of professionalism as espoused by Huntington, what was really needed to keep the military out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the concept are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed in debates about civil-military relations. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific who is also well known for the controversy surrounding his conduct in the Korean War and his eventual dismissal by President Truman, told the cadets at West Point in May of 1962.

@@

Current military doctrine and practices likewise reinforce the notion of the military remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. General Martin Dempsey, in an opinion piece published in the Washington Post on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although the notion of an apolitical military is the theoretical goal of the institution, the practical record remains far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory from time to time. What we have established, though, is clear: a non-partisan military that performs its military mission well has long been the goal of both civil-military relations theorists and military officers. More importantly, this is what fosters and ultimately earns the trust and confidence of the American people.

## Political Polarization, Public Opinion, and Military Expertise

Scholars of American Politics have studied political polarization in depth, and while they have offered differing explanations for the causes of polarization, ranging from race and religion [@abramowitz_great_2018], to debates about the size and role of government [@sides_two_2015], to fierce policy disagreements over policy [@mann_its_2012], virtually all agree that polarization, if not an entirely new phenomenon, has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014]. Hare and Poole are among scholars who argue that modern American political polarization began in the 1970s as a result of redistribution along racial lines following key pieces of legislation in the 1960s [@hare_polarization_2014]. They further argue that polarization disagreement has only intensified, something that a significant portion of modern Americans agree on [@schaeffer_far_nodate].

Within this polarized political environment, what role does the military play - particularly retired military officers who possess considerable military expertise? In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argued that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information. He is rather acknowledging that most citizens care most about the world that is not "beyond their personal experience" [@zaller_nature_1992]. His point is that elites play a significant role in helping to form and sway public opinion.

Michael Robinson further examined how elite cues are received and processed. Robinson argues that partisan groups, as a result of ongoing political polarization, are cognitively biased in distinct ways in how they receive and process information [@robinson_who_2019]. As a result, Robinson found that partisans will express confidence in the military in vastly differing ways even after neutral, non-partisan information about various types of military failure is presented to these groups [@robinson_who_2019]. If Republicans are less likely to express a reduction in confidence in the military as an institution than Democrats are, as Robinson alleges, then even accurate, unbiased information is potentially irrelevant, as these differing partisan groups are pre-programmed to selectively see a particular situation as they wish to see it [@robinson_who_2019]. His main point serves to warn us regarding confidence in the military as an institution: there may be real reasons to lose confidence in the military as a result of a particular incident, but partisan filters will prevent a reduction of confidence from actually being expressed.

Zaller's framework makes it clear that it is a perfectly reasonable assumption that retired military officers who choose to express their opinions through the form of opinion-editorial pieces constitutes a form of elite discourse. After all, retired military officers have at least twenty years of experience, and in the case of most multi-star generals or admirals, usually more than 30 and in some cases close to 40 years of military experience. Indeed, most retired senior generals and admirals have spent their professional careers solely in the military, and have extensive experience while serving in leadership positions of large organizations and, in many cases, during times of war. Retired military officers are thus elites in the field of national security, and in line with Zaller's framework, they possess the potential to significantly contribute to policy debates and thus, to influence public opinion. Yet we also know that the American public will not necessarily evaluate the opinions of these military experts objectively, but rather through pre-conceived and potentially partisan lenses [@robinson_who_2019].

# Methodological Approach, Data, and Hypotheses

Given these three parameters -- that retired military officers are part of an institution that highly values professionalism and non-partisanship, that political polarization in America has increased since at least the 1970s, and that retired military officers are elites who possess the potential to sway public opinion -- I now aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened. To answer this question, I assemble and examine a data set that consists of opinion pieces authored by retired military officers in major US newspapers. My methodological approach involves two key steps. First, I determine if there have been shifts in the topics that retired military officers have been writing about. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed. In simple terms, I want to answer the questions, what are these retired military officers writing about, and through the lens of professionalism, should they be saying what they are saying?

Using the Factiva and Nexus-Uni databases, I conduct a search that focuses keywords in the byline (author field) such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth. For sources, I confine my search to the largest US newspapers that regularly run opinion-editorial pieces, which includes  _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. I begin the search for the data of January 1, 1979 - 30 June 2020^[I also open my search to include both print and online offerings to ensure that I am capturing any opinion pieces published digitally. Most of the newspapers initiated digital offerings in the late 1990s.]. Using newspapers, and particularly these specific ones, provides some advantages as well as disadvantages. The primary advantage is that these sources have been around since the period I wish to examine, unlike many of the online only sites that have only been present in the digital age. This allows me to form a data set which covers more time. The primary disadvantage is that I am not capturing all of the opinion-pieces authored or given by retired military officers. For example, retired military officers are common on cable TV, often weighing in on topics, and I am not capturing those here. It is also true that over time,  social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, but overall circulation is still down today compared to earlier years, even after factoring in the expanse of digital options available^[As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million and 54 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million [@barthel_leading_2019]]. Nonetheless, I contend that the data I am able to assemble will still be informative. Publishing opinion-editorial pieces, especially in major, competitive newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by figures that include heads of state, members of congress, governmental figures, and leaders of industry and academia. Additionally, opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. They are different from interviews or commentary that may be asked by a questioner because the author must take time to put forth his or her position in writing. The data therefore allows the researcher to truly ascertain what the specific views and arguments of an author were on a given topic or policy.

Of note, when assembling the data, some authors are explicitly identified in some of their publications as retired military officers, but not in others.^[This was true of several individuals, including Alexander Haig, a retired Army General who later served as Secretary of State under Ronald Reagan;  Robert McFarlane, a retired Marine Corps Lieutenant Colonel who later served as the National Security Adviser to President Reagan, Andrew Bacevich, a retired Army Colonel who then became a prominent historian and author; and Michael Hayden, a retired Air Force General who served as director of the National Security Agency and the Central Intelligence Agency]. I conduct a separate search for all opinion pieces penned by these authors, regardless of whether they were identified as a retired military officer or not, concluding that it is best do so for coding purposes. Ultimately, the data set contains 334 opinion-editorial publications, penned by 178 different authors or author teams.^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] 10 of these pieces are written unanimously, with nods by the authors to their previous service.

From a macro perspective, consider for a moment all that the date range includes with respect to the US military. Indeed, numerous and substantive events, debates, and changes have occurred in this time period for retired military officers to weigh in on. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, as well as a host of smaller but important military operations in Grenada, Panama, Somalia, and Haiti. The military has also undergone significant social and institutional changes in the time period examined. In the late 1970s, the military had ended the draft only years before following the conclusion of the Vietnam War. The debates over "Don't Ask, Don't Tell" and policies regarding women in the combat arms are also included in the data, as are numerous discussions of military budgets, weapons systems, and alliances. 
Attempting to reconcile the fact that the military as an institution prides itself on being a non-partisan and apolitical institution, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 40 years, and that the US political environment is increasingly polarized along partisan lines, I adopt the following hypotheses:

**$H_1$: The majority of op-eds published by retired military officers over the time period examine will be apolitical and non-partisan in tone.**

**$H_2$: Because overall newspaper consumption has dropped in the past 40 years, there will be fewer opinion pieces over time authored by military officers.**

**$H_3$: Given that polarization has become increasingly pronounced in the United States in recent years, there will be an increase in the number of op-eds which reveal political and partisan sentiments (indication of a drop in professionalism).**

# First Layer: Topical Analysis

In the first level of analysis, after capturing key biographical and publication information, such as author name, article title, source, publication date, and branch of service and rank of the author, I code each observation into one of several binary categories that corresponds to the overall subject matter of the publication. The goal is to create a coding scheme by subject that is sufficiently broad but also specific enough that is helpful in differentiating the overall main point of each publication. The following coding scheme is therefore devised and implemented (all variables are binary, taking on the value of 1 if the publication corresponds to the general definition of the variable, and 0 otherwise): \singlespace

* **War_Ops**- if the piece advocates for or denounces a particular course of action directly related to operational or war fighting decisions of a clear military nature. Examples include: the wisdom of troop increases, the Surge in Iraq, whether to bomb Serbia, or even the role of the prison at Guantanamo.

* **Trp_Clt**- if the piece generally showcases or highlights the work and sacrifice of members of the armed forces; to defend the work of specific members of the Armed Forces; or to call for policing up the institutional cultures of the services. Examples include: remembering our fallen soldiers on Memorial Day; improving the institutional culture of the Army or Navy; advocating on behalf of or remembering a specific flag officer.

* **Str_Adv**- if the piece is generally written to offer some sort of strategic advice for the nation, to include the wisdom of pursuing particular alliances or agreements, or adopting a specific weapons platform or military budget. Examples include: why the US needs to strengthen NATO, build ties with China, increase defense spending, or not retire the A-10 combat plane.

* **MilPSoc**- if the piece advocates for or denounces a particular social policy that _directly influences_ the armed forces. Examples include: debates over keeping or removing the "Don't Ask, Don't Tell" policy, opening up combat roles to women, the role of Transgender members of the armed forces.
